Universals of word order result from optimization of grammars for efficient communication

### 2019-02-20 second-draft/3-paper-names.tex
new draft

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.

In this work we examine this question from a computational perspective, demonstrating a fully formalized framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

We aim to explain the linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. These universals are called \key{implicational} universals, and take the form of correlations between certain word order patterns and others across languages. For example, whether a language puts verbs before or after objects strongly determines whether it places genitives before or after nouns, and whether it has prepositions or postpositions \cite{dryer1992greenbergian}.

A prominent line of research has argued that universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

Researchers working in the functional paradigm have proposed a range of criteria which languages should meet in order to enable efficient communication and processing. These criteria are based on theories of online processing difficulty (see \cite{jaeger2011language} for a review) and on information theoretic notions of efficiency and robustness in communication \cite{ferrericancho2001two,piantadosi2011word,gibson2013noisy}, and they have been formalized to varying degrees.

We formalize the notion of communicative efficiency using two terms: one having to do with minimizing the difficulty of linguistic communication, and one having to do with maximizing the information transferred in such communication. We quantify the difficulty of linguistic communication in terms of the word-by-word \key{predictability} of strings, which has been found to be a highly accurate and general predictor of human online comprehension difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}, and which can be justified in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}. We quantify the information transferred in language as \key{parseability}: the quantity of information contained in linguistic utterances about latent hierarchical tree structures. We also compare the results of optimizing for communicative efficiency to the results of optimizing for dependency length minimization, a previously proposed approximate metric of the amount of memory involved in language production and comprehension \cite{hudson1995measuring,gibson1998linguistic,gibson2000dependency,liu2017dependency,temperley2018minimizing}.

%To evaluate whether communicative efficiency can explain word order universals, we apply the following method \cite{gildea2007optimizing}. First, we develop word order grammars: simple, interpretable models of word order, implemented as functions that take in an unordered syntactic dependency tree and output a distribution over possible word orders for that tree. Second, we take dependency treebanks of natural languages and compute word order grammars which are optimized for an objective functions representing communicative efficiency.
%Third, we evaluate whether the optimized word order grammars reproduce the desired word order universals when applied to attested dependency trees from the treebanks across many languages. Our favorable results suggest that it may be possible to generally model human languages as efficient codes for communication among agents with generic human-like cognitive and communicative constraints. %The best way to describe a human language might be as a solution to an equation involving such constraints.

\section{Greenberg's Word Order Correlations}
Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
`To' is expressed by a \emph{preposition}, so named because it precedes the word expressing `friend'. %The adpositional phrase starts with the adposition.
In Japanese, the verb stands at the end; object and adpositional phrase precede it; `to' is expressed by a \textit{postposition}, so named because it follows the word denoting `friend'.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb, and that the objects of adpositions and verbs stand on the sam side of their heads.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
%It turns out that these are deep and stable regularities:
It turns out that this variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions:
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.
%However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
This correlation has an extremely strong empirical basis:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

TODO maybe being more general in the beginning, as in the first version. make sure the reader doesn't think 'it's just headedness, why bother'

@@

Zipf \cite{zipf1949human} argued that language optimizes a tradeoff between speaker effort (Force of Unification) and listener effort (Force of Diversification).
According to this theory, a language that minimizes speaker effort should reduce the number of different utterances to make production as easy as possible.
Conversely, to minimize listener effort, the language should provide different utterances for different meanings, so that the listeenr can unambiguously identify the meaning from the utterance.
Minimizing speaker and listener effort represent two opposing forces:
A pressure to reduce the complexity of the language makes the utterances in a language more homogenous.
A pressure to reduce ambiguity makes utterances more heterogeneous, so they can indicate different meanings.
%Another way to conceptualize these two pressures is as the \key{cost} and \key{informativity} of language use:
Producing and processing simple utterances incurs little cost, but more complex and diverse utterances are required to be provide higher informativity.
The idea that language results from the tension between these two pressures has a long and fruitful history.
\cite{horn-toward-1984}, RSA, Color naming

%In the domain of word order,

%TODO introduce the terms \key{cost} and \key{informativity}
%
This idea has been shown to account for phenomena such as pragmatic inference \cite{horn-toward-1984, frank2012predicting} and color naming~\cite{zaslavsky2018efficient}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/grammars.png}
	\caption{Given a set of dependency structures, grammars define a set of sentences. For simplicity, we use English words. Each grammar defines a consistent ordering for the different kinds of syntactic relations, e.g., Grammars 1 and 2 order the object after the verb, Grammars 3 and 4 order the object before the verb. }
	\label{fig:grammars}
\end{figure}

>>>>

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. In this work we examine this question from a computational perspective, demonstrating a fully formalized framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

\textbf{TODO maybe should be more general in the beginning, as in the first version. Also make sure the reader doesn't think 'it's just headedness, why bother'.}

Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
`To' is expressed by a \emph{preposition}, so named because it precedes the word expressing `friend'. %The adpositional phrase starts with the adposition.
In Japanese, the verb stands at the end; object and adpositional phrase precede it; `to' is expressed by a \textit{postposition}, so named because it follows the word denoting `friend'.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb, and that the objects of adpositions and verbs stand on the sam side of their heads.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
%It turns out that these are deep and stable regularities:
It turns out that this variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions:
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.
%However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
This correlation has an extremely strong empirical basis:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

@@

A commonly proposed functional explanation of Greenberg's correlations is in terms of Dependency Length Minimization \cite{futrell2015largescale,liu2017dependency,temperley2018minimizing}.
This is the idea that languages order words so as to minimize the linear distance between words linked in head--dependent relationships.
This has been confirmed on the basis of large-scale corpus evidence \cite{futrell2015largescale,liu2017dependency,temperley2018minimizing}, and has been justified based on the reasoning that long dependencies create high memory requirements for parsing and generation \cite{gibson1998linguistic,gibson2000dependency}.

%In previous studies of word order optimization, it has been common to propose that word orders are chosen to minimize \key{dependency length}, the linear distance between words linked in head--dependent relationships \cite{futrell2015largescale,liu2017dependency,temperley2018minimizing}. 
%Dependency length has been proposed as a metric of processing difficulty on the reasoning that long dependencies create high memory requirements for parsing and generation \cite{gibson1998linguistic,gibson2000dependency}, and minimization of dependency length is known to recover a large subset of the Greenbergian universals \cite{hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic}. 
%We demonstrate that the minimization of dependency length can be formalized in our framework, reproducing the finding that optimizing grammars for dependency length recovers several word order universals \cite{gildea2010grammars}.

Zipf \cite{zipf1949human} argued that language optimizes a tradeoff between speaker effort (Force of Unification) and listener effort (Force of Diversification).
According to this theory, a language that minimizes speaker effort should reduce the number of different utterances to make production as easy as possible.
Conversely, to minimize listener effort, the language should provide different utterances for different meanings, so that the listeenr can unambiguously identify the meaning from the utterance.
Minimizing speaker and listener effort represent two opposing forces:
A pressure to reduce the complexity of the language makes the utterances in a language more homogenous.
A pressure to reduce ambiguity makes utterances more heterogeneous, so they can indicate different meanings.
%Another way to conceptualize these two pressures is as the \key{cost} and \key{informativity} of language use:
Producing and processing simple utterances incurs little cost, but more complex and diverse utterances are required to be provide higher informativity.
The idea that language results from the tension between these two pressures has a long and fruitful history.
%\cite{horn-toward-1984}, RSA, Color naming
%
%In the domain of word order, 
%
%TODO introduce the terms \key{cost} and \key{informativity}
%
This idea has been shown to account for phenomena such as pragmatic inference \cite{horn-toward-1984, frank2012predicting} and color naming~\cite{zaslavsky2018efficient}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/grammars.png}
	\caption{Given a set of dependency structures, grammars define a set of sentences. For simplicity, we use English words. Each grammar defines a consistent ordering for the different kinds of syntactic relations, e.g., Grammars 1 and 2 order the object after the verb, Grammars 3 and 4 order the object before the verb. \textbf{TODO This figure is just a draft sketch so far.}}
	\label{fig:grammars}
\end{figure}
